Full simulation

### 2024-06-11 full_simulation.qmd
update bibfile, add full_simulation.qmd notebook, change notebook-links option

---
title: "A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models"
shorttitle: "Simulation-Based Sample Size Planning for GLMMs"
author:
  - name: Florian Pargent
    corresponding: true
    equal-contributor: true
    orcid: 0000-0002-2388-553X
    roles:
      - formal analysis
      - methodology
      - visualization
      - writing
      - editing
    affiliations:
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
  - name: Timo K. Koch
    corresponding: true
    equal-contributor: true
    orcid: 0000-0001-6728-2063
    email: timo.koch@unisg.ch
    roles:
      - conceptualization
      - formal analysis
      - writing
      - editing
    affiliations:
      - name: University of St. Gallen 
        department: Institute of Behavioral Science & Technology
        address: Torstrasse 25
        city: St. Gallen
        country: Switzerland
        postal-code: 9000
      - ref: lmu
  - name: Anne-Kathrin Kleine
    roles:
      - formal analysis
      - editing
    affiliations:
      - ref: lmu
  - name: Eva Lermer
    roles:
      - conceptualization
      - funding aquisition
      - editing
    affiliations:
      - ref: lmu
      - name: Technical University of Applied Sciences Augsburg
        department: Department of Business Psychology
  - name: Susanne Gaube
    roles:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: lmu
      - name: University College London
        department: Global Business School for Health
author-note:
  disclosures:
    data-sharing: "All materials (reproducible manuscript incl. R code, simulation results) are available in the project's repository on the Open Science Framework (OSF) at https://osf.io/dhwf4/"
    conflict-of-interest: "The authors declare that there were no conflicts of interest with respect to the authorship or the publication of this article."
    financial-support: "This research was funded by a grant from the Volkswagen Foundation (Grant No. 98525)."
    gratitude: ~
    authorship-agreements: "Florian Pargent and Timo K. Koch contributed equally to this work."
abstract: "When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. The recent replication crisis underlines the need for more rigorous statistical methodology and adequately powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. However, available methods and software for a priori sample size planning for GLMMs are often limited to specific designs. Tailored data simulation approaches are a more flexible alternative. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori sample size planning with GLMMs. We not only focus on power analysis but also explain how to use the precision of parameter estimates to determine appropriate sample sizes. We conclude with an outlook on the increasing importance of simulation-based sample size planning."
keywords: [power analysis, precision, data simulation, sample size planning, generalized linear mixed model]
bibliography: [Simulation Tutorial Paper.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

```{r}
#| label: setup
#| include: false
library("papaja")
library("tidyverse")
library("faux")
library("lme4")
library("multcomp")
library("marginaleffects")
library("future")
library("furrr")
library("binom")
r_refs("r-references.bib")
```

# Introduction

When planning experimental research, it is essential to determine an appropriate sample size and to use appropriate statistical models to analyze the data to ensure that the results obtained are both robust and informative [@lakensSampleSizeJustification2022]. The recent replication crisis has illustrated many challenges surrounding the reproducibility and reliability of study findings [@yarkoniGeneralizabilityCrisis2022]. As a result, there is a growing need for more rigorous statistical methodology and the adoption of adequately powered experimental designs. Multiple easy-to-use software solutions exist for simple statistical models and experimental designs. However, many researchers lack the skills and tools to conduct "a priori" (i.e., before data collection) sample size planning for more complex research designs such as flexible generalized linear mixed models (GLMM) framework. In the present work, we provide a tutorial on how to determine adequate sample sizes by performing tailored simulation-based sample size planning for GLMMs. After introducing some theoretical background on sample size planning, we review existing software solutions in R and discuss under which circumstances tailored data simulations are necessary. Then we describe general steps and decisions involved in tailored data simulation. To illustrate the details of these steps, we finish with a hypothetical case study from the field of human-AI (artificial intelligence) interaction research.

<!-- # Recommended background knowledge -->

To benefit most of this tutorial paper, we recommend readers to familiarise themselves with basic statistical concepts like hypothesis tests (HTs) and their statistical power as well as confidence intervals (CIs) and their precision [@lakensImprovingYourStatistical2022; @riesthuisSimulationBasedPowerAnalyses2024; @kumleEstimatingPowerGeneralized2021]. Some knowledge of causal inference is useful but not necessary [@lundbergWhatYourEstimand2021; @deffnerCausalFrameworkCrossCultural2022]. In addition, readers should have a rough understanding of R [@wickhamDataScienceImport2023] and how to simulate data. For data simulation, we use functions from the tidyverse [@wickhamWelcomeTidyverse2019] and the faux package [@R-faux].
Finally, readers should be familiar with regression modeling in general and GLMMs in particular. In this tutorial, we simulate data by manually specifying the model equation of a GLMM that represents our assumed data generating process [@debruineUnderstandingMixedEffectsModels2021]. It is not necessary to understand the technical details of how GLMMs are estimated. However it is crucial to understand the structure of a basic GLMM (e.g., logistic regression with random intercepts) and how the model assumes that the values in the dependent variable are determined by the predictor variables and the random effects.

# Theoretical background

## Planning for statistical power or precision

Conducting research with insufficiently large sample sizes can have many negative consequences. First, experiments may yield inconclusive or misleading results, hindering the accumulation of knowledge. Second, studies that are doomed to never finding a postulated effect waste resources by consuming time, effort, and funding without delivering meaningful results. For these reasons, many journals and funding bodies now require that a justification of sample size is included in study protocols and grant proposals, recognizing its significance in ensuring robust and meaningful findings.
Although scientists often do not justify sample size or use general heuristics from the literature, resource constraints often play an important role [@lakensSampleSizeJustification2022].
But ideally, a suitable sample size should be determined a priori (i.e., before the study is conducted) based on some meaningful computation to ensure that the study will be able to fulfill its purpose.

The majority of empirical studies in psychology and other social sciences apply hypothesis testing. As a consequence, the dominant approach for determining an adequate sample size is based on power analysis (i.e., planning for power) [@lakensSampleSizeJustification2022; @maxwellSampleSizePlanning2008]. 
Statistical power is defined as the probability that a HT has a significant p-value when analyzing repeated samples from a population with a true effect of some pre-specified size [@cohenPowerPrimer1992]. Less formally, power is described as the probability that a HT correctly rejects the null hypothesis when the alternative hypothesis is true. If the sample size (i.e., the number of participants and/or stimuli) used for data collection is insufficient to detect the effects or relationships being investigated with high probability, the study is considered "underpowered".
When planning for power, a target is set for the statistical power of a HT of interest. Assuming some effect size of interest and a desired significance level, a minimum sample size can be determined that, on average, would guarantee reaching this target. Although it is recommended to justify the desired significance level and power [@lakensJustifyYourAlpha2018], most empirical studies adopt the heuristic of $\alpha = 0.05$ and $1 - \beta = 0.80$.

In contrast to power analysis, sample size planning can also be based on the precision of parameter estimates (i.e., planning for precision or planning for accuracy) [@lakensSampleSizeJustification2022; @maxwellSampleSizePlanning2008]. Not all research questions are best answered using hypothesis testing. It has been argued that basic research rarely requires making discrete decisions on whether some effect has been "discovered" and should thus shift from a hypothesis testing towards an estimation framework [@mcelreathStatisticalRethinkingBayesian2020; @cummingNewStatisticsWhy2014; @kruschkeBayesianNewStatistics2018]. Although this view is not without critique [@uyguntuncEpistemicPragmaticFunction2023], at least for exploratory or pilot studies where little previous research has been conducted, more scientists seem to agree that simply estimating the effects of interest and making the estimation uncertainty transparent by reporting CIs is more useful. Assuming that no HTs are conducted for a planned study, power analysis is not relevant for sample size planning. Nonetheless, the sample size still has a crucial effect on how informative the planned study will be, because an effect of interest is estimated more precisely with bigger samples. In the precision framework, the target quantity commonly used for sample size planning is the expected width of a CI [@kelleySampleSizePlanning2006; @lakensSampleSizeJustification2022; @maxwellSampleSizePlanning2008]. The values inside a CI are often interpreted as plausible values for the quantity of interest it is supposed to estimate. More formally, a CI with confidence level 0.95 provides the smallest interval with the property that upon repeated sampling, 95% of individual CIs would include the true quantity of interest. Thus, a narrow CI is more informative about the size of the true effect than a wide interval. Apart from the desired confidence level, the width of a CI depends strongly on the sample size. Because bigger samples carry more information, they lead to smaller CIs. When planning for precision, a target can be set for the expected width of a CI of interest. Assuming some effect size of interest and a desired confidence level, a minimum sample size can be determined that, on average, would guarantee reaching this target. Because planning for precision is still rare, there are no common heuristics on how to choose the desired width of the CI [@lakensSampleSizeJustification2022].

## Generalized linear mixed models (GLMMs)

As study designs become more complex, psychological researchers require more sophisticated statistical models to capture the nuanced relationships and grouping structures introduced by them [@yarkoniGeneralizabilityCrisis2022]. GLMMs (also called multilevel models) are gaining increasing popularity because they offer great flexibility [@fahrmeirRegressionModelsMethods2021].
GLMMs are an extension of LMMs (Linear Mixed Models), which are, in turn, extensions of linear regression models that account for correlated data, including hierarchical structures [@fahrmeirRegressionModelsMethods2021]. In this context, correlated data means that the value in the outcome variable for one observation may be related to the value of another observation in a systematic way that is not already accounted for by the usual (fixed) predictor variables (e.g., age of participants). This correlation can arise for various reasons: For instance, responses to some stimuli from some participants might be more similar because the same person was measured twice (repeated measurements), participants come from the same neighborhood (clustering), or participants responded to the same stimulus (stimulus effects). Thus, modeling such correlations is important whenever the data has a clear structure, while the grouping variables can be hierarchically nested (e.g., grouping variables students and schools: each student belongs to exactly one school) or cross-classified (e.g., grouping variables students and math exercises: each student is presented with several math exercises). LMMs are used when the outcome variable is continuous and follows a normal distribution. They allow for the modeling of fixed effects, which capture the relationships between the usual predictors and the outcome, as well as random effects, which account for the different types of correlation structure and grouping effects. Random effects are typically assumed to follow a normal distribution with a mean of zero and a variance that quantifies the heterogeneity across groups.
GLMMs extend the LMM framework to accommodate non-normally distributed continuous and categorical outcome variables. GLMMs involve a link function that connects the linear combination of predictor variables to the expected value of the outcome variable. The link function allows for modeling the relationship between predictors and the outcome in a non-linear way that is appropriate for the specific distribution family of the outcome variable. For example, think of an experiment with different design factors (e.g., picture positions, headline aesthetics) impacting the likelihood of users clicking on an online advertisement. The participants’ behavior is measured repeatedly over several sessions. The click patterns of participants in one session are likely to be correlated with their previous sessions and the outcome variable is binary (click/no click) for each session, which follows a binomial distribution.

# Simulation-based sample size planning with GLMMs

<!-- For less complex statistical models, like t-tests, ANOVA, and linear regression, with common study designs (e.g., mean comparison between two groups), user-friendly software for a priori sample size planning is readily available [@faulStatisticalPowerAnalyses2009; @champelyPackagePwr2018; @kelleyObtainingPowerObtaining2003; @kelleySampleSizePlanning2006].  -->
<!-- However, these software packages are often not flexible enough to perform sample size planning for complex designs. -->

To our knowledge, existing approaches for sample size planning for GLMMs have exclusively focused on planning for power. Power analysis methods for multilevel models can be categorized into formula-based methods and simulation-based methods [@murayamaSummarystatisticsbasedPowerAnalysis2022]. Formula-based methods rely on formulas to calculate power directly while simulation-based methods rely on repeatedly simulating data with a known true effect size and estimating power empirically (i.e., what percentage of simulated datasets produces a significant p-value). Available formula-based software packages for multilevel models often do not include GLMMs or are limited to very simple designs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022], making it necessary to build data simulations tailored specifically to the study design. A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021; @kainPracticalGuidePower2015; @greenSIMRPackagePower2016a]. However, many of these tutorials focus on linear mixed models (LMMs) and the common study designs [but see @kumleEstimatingPowerGeneralized2021, for a tutorial that also covers more advanced settings]. This narrow focus provides limited guidance for researchers faced with more complex study designs, especially when little prior knowledge about plausible effect sizes is available [see the discussion in @kumleEstimatingPowerGeneralized2021]. Simulation-based power analysis with GLMMs requires making a range of assumptions: The (conditional) distribution assumption specifies the distributional family for the outcome variable. Assumptions about the random effects include the assumption of normality (i.e., that the random effects follow a normal distribution) and the covariance structure among the random effects (i.e., if and how they are correlated). Making these decisions requires understanding the underlying assumptions of the model and ensuring they align with the characteristics of the data being analyzed. Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions a practical challenge (see the discussion in [@maxwellSampleSizePlanning2008] and [@kumleEstimatingPowerGeneralized2021]).

In Table XXX, we give a short review of existing R packages that can be used for power analysis for GLMMs.

# When to use tailored data simulation?

Performing tailored simulation-based sample size planning is more complicated and time-consuming than using the existing software tools outlined in TABLE XXX. The most important circumstances under which tailored simulation-based sample size planning is necessary are 1) complex study designs, 2) complex statistical hypotheses, 3) planning for precision, 4) no available prior studies or pilot data.

<!-- ## Complex study designs -->

Requirements of real-world studies are often more complex than the simplified designs assumed by many user-friendly software packages for sample size planning.
One frequent issue in applied data analysis is missing data, and there can be various reasons for this [@littleStatisticalAnalysisMissing2014].
For example, data can be missing completely at random (e.g., because an electronic measurement device randomly failed for some technical reasons).
Alternatively, subjects might systematically drop out or produce missing data, but this dropout can be explained by some attributes also measured in the dataset (e.g., older subjects have a higher probability to refuse answering a question on income).
In a more complicated scenario, missing data in some variable is caused by the measured attribute itself (e.g., wealthy people are more likely to refuse reporting their income).
Moreover, many experimental designs contain conditions in which values of the predictor variables are missing by design. This can make data analysis more complicated because predictors have to be coded in specific ways that prevent the estimated GLMM from becoming unidentified. Whether missing data has an effect on the sample size planning depends on our theoretical assumptions on how the missingness is caused.
However, it is often challenging to decide whether missing data can be safely ignored in the data analysis and sample size planning process based on a merely theoretical approach [@gomilaMissingDataExperiments2022].
Tailored simulation-based approaches offer the possibility to include the assumed process of how data become missing in the data simulation, thereby determining the required sample based on simulated datasets that contain missing values [for an example, see @lanePowerStrugglesEstimating2018]. As a byproduct, the simulated datasets can also be used to test whether the intended data analysis provides the expected (unbiased) results, despite the missing data.
Although GLMMs can handle a large variety of outcome variables, researchers are becoming increasingly aware that many datasets might profit from even more sophisticated models. Common examples are zero-inflated outcomes, censoring, and nonlinear predictor effects that can be modeled with the R packages glmmTMB [@brooksGlmmTMBBalancesSpeed2017] or brms [@burknerAdvancedBayesianMultilevel2018]. 
Tailored simulation-based approaches do not share the same limitation than the existing software solutions for power analysis that focus exclusively on GLMMs.
As long as there is a software package available to estimate the model of interest, it is always possible to perform tailored simulation-based sample size planning.

<!-- ## Complex statistical hypotheses -->

The most common hypotheses tested in psychological research are of the type $H_0: \beta = 0$, where $\beta$ is a slope or intercept of a regression model.
However, many research questions in psychology actually require testing more complex statistical hypotheses.
In the new era of preregistration and registered reports [@chambersPresentFutureRegistered2022], most research questions should be tested with directed hypotheses because good theories at least postulate whether some psychological effect of interest is positive or negative. Even better theories should be able to specify the smallest effect sizes of interest (SESOI) that must be exceeded if the effect has any practical relevance [@lakensEquivalenceTestingPsychological2018]. In combination, this might require a test such as $H_0: \beta \leq 0.1$.
More elaborate research questions often require testing hypotheses that consist of a combination of model parameters, for example testing simple slopes [@preacherComputationalToolsProbing2006] with a hypothesis such as $H_0: \beta_0 + \beta_1 \leq 0$.
If the research question consists only of a single hypothesis of this sort, it might be possible to reduce the hypothesis to a single regression coefficient by clever coding and/or centering of predictor variables. However, interesting research questions often consist of combined hypotheses that consist of more than one separate statistical hypothesis [for a tutorial on contrast analysis in GLMMs, see @schadHowCapitalizePriori2020]. For example, a combined null hypothesis $H_0$ might consist of two single null hypotheses $H_{01}: \beta_1 \leq 0$ and $H_{02}: \beta_0 + \beta_1 \leq 0$.
For some research questions, the combined null hypothesis $H_0$ would be rejected if both $H_{01}$ *AND* $H_{02}$ are rejected.
For other research questions, the combined null hypothesis $H_0$ would be rejected if $H_{01}$ *OR* $H_{02}$ *OR* both are rejected.
If the global hypothesis $H_0$ is combined with *OR*, the p-values of the single hypotheses must be corrected for multiple testing to avoid $\alpha$-inflation for the global hypothesis [@dmitrienkoTraditionalMultiplicityAdjustment2013].
However, if the global hypothesis $H_0$ is combined with *AND*, a correction for multiple testing is not necessary but rather a mistake that unnecessarily reduces the power of the global HT.
None of the software packages for sample size planning in table X can handle combined hypotheses as discussed here and only some can handle directed hypotheses. In contrast, our case study will demonstrate how we can test directed combined hypotheses with tailored simulation-based sample size planning.

<!-- ## Planning for precision -->

Although planning for precision [@lakensSampleSizeJustification2022; @maxwellSampleSizePlanning2008; @cummingNewStatisticsWhy2014] has been increasingly discussed as a useful strategy for empirical research, all available software packages for sample size planning with GLMMs are based on power analysis.
Therefore, researchers that want to want to apply an estimation strategy in their studies, without testing any statistical hypotheses, currently cannot use the software packages outlined in TABLE XXX.
<!-- As long as one is interested only in confidence intervals of a single model parameter, it might be possible to adapt existing software packages to plan for precision. However, we demonstrate in our case study that in realistic research projects with GLMMs, the contrasts of interests and their respective confidence intervals are usually situated on the scale of the (expected) outcome in contrast to the scale of the model parameters.  -->
However, tailored simulation-based sample size planning can easily handle the planning for precision approach [@maxwellSampleSizePlanning2008]. The only change in procedure is that instead of computing HTs for each simulated dataset and estimating statistical power across repetitions, CIs are computed for each simulated dataset, and the expected width is estimated.

<!-- ## No prior studies or pilot data -->

All frameworks for sample size planning require the user to make assumptions about the expected effect size. Assuming the true effect is of this size (or greater), one can compute the (minimum) power of a HT or the (maximum) expected width of a CI. Existing software packages for sample size planning for GLMMs usually require to provide the assumed effect in the unit of some standardized measure of effect size. When the researcher has access to similar studies or pilot data, providing such standardized effect sizes is feasible. However, note that choosing effect sizes based on small pilot studies is generally not recommended, as those estimates can be heavily biased [@albersWhenPowerAnalyses2018a; @lakensSampleSizeJustification2022].
Providing an informed standardized effect size can be an almost impossible challenge when no prior studies of pilot data are available. This problem is further exacerbated by the fact that GLMMs are so flexible that general heuristics of what should be considered a small effect do not exist or are difficult to defend. In our experience, using domain knowledge to construct a tailored data simulation is the only solution to determine plausible effect sizes in the absence of prior evidence. It would be possible to use these tailored simulations to extract plausible values for standardized effect sizes that could then be inserted in existing software packages for sample size planning. However, we would argue that when tailored data simulations necessary to determine effect sizes anyway, performing the whole sample size planning in a customized way is preferred over using the existing software packages.

# General steps in tailored simulation-based sample size planning

Although the details differ depending on the specific study characteristics, each tailored simulation-based sample size planning requires a series of steps and decisions. We will introduce each step in a theoretical section, followed by the practical application based on a hypothetical case study. All code in this manuscript and simulation results are available in the project’s repository on the Open Science Framework (https://osf.io/dhwf4/).

## Step 1: Define the estimand

### THEORY

The first step in every research process should be a clear definition of the theoretical *estimand* [@lundbergWhatYourEstimand2021], i.e. the theoretical quantity which is necessary to answer a specific research question. The estimand consists of a quantity, that can be described for each unit under investigation and a clear definition of the target population, for which the quantity is of interest. 
For example, an estimand might be the probability that a clinical psychologists makes the correct diagnosis for a psychiatric patient with major depression, averaged across all clinical psychologists and depressed patients in psychiatric institutions in Germany.

The estimand should always be defined outside of any statistical model, because there are usually a range of statistical methods that can be used to estimate the same estimand, depending on the study design (e.g., a randomized experiment) that will produce the observed data in the planned study.
For many common research questions in psychology, the estimand can be expressed as a statistical quantity that can be estimated with a regression model, for example a single $\beta$ coefficient. However, this is not possible for all estimands, which is why the literature discusses many estimation strategies beyond regression [@lundbergWhatYourEstimand2021; @deffnerCausalFrameworkCrossCultural2022].

<!-- When the empirical estimand has been specified, the last step is to define how a concrete estimate of the empirical estimand shall be computed based on the observed study data. When a regression model is used for estimation, this includes decisions like which statistical framework estimation technique and software package are used to fit the model and how to compute HTs or CIs. -->

### PRACTICE

In the present hypothetical case study, we consider the effectiveness of feedback provided by an artificial intelligence (AI) embedded in a diagnostic decision support system. The context is a clinical setting, where expert radiologists and students under training must detect bleeding based on head scans from computer tomography (CT). In the investigated AI-enabled diagnostic decision support system, an AI model can provide initial diagnostic advice, which can be used as guidance by the humans who are required to make the final diagnostic decision. The research goal is to validate the effectiveness of the AI-enabled advice. We consider the AI-enabled advice as effective, if the following pattern holds, which we will first describe verbally:

*We expect that for BOTH expert radiologists and medical students, correct AI-advice leads to a higher probability of accurately diagnosing a CT scan compared to no AI-advice presented, AND, we expect that for BOTH experts and non-experts, incorrect advice leads to a lower probability of accurately diagnosing a CT scan compared to no advice presented.*

It becomes clear that our estimand consists of four comparisons between experimental conditions [@lundbergWhatYourEstimand2021].
However, the verbal description is still somewhat vague, which is why we try to give a more precise expression for each comparison:

$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{correct advice, average expert, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{no advice, average expert, average scan})
\end{aligned}
$$

$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{no advice, average expert, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{wrong advice, average expert, average scan})
\end{aligned}
$$
$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{correct advice, average student, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{no advice, average student, average scan})
\end{aligned}
$$

$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{no advice, average student, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{wrong advice, average student, average scan})
\end{aligned}
$$
For example, the first expression is concerned with the difference between the probability that a correct diagnosis is made if correct AI-advice is presented and the probability that a correct diagnosis is made if no AI-advice is presented. This contrast is quantified for a hypothetical *typical* expert and a *typical* scan, where *typical* is usually defined as an average score on all attributes of the expert or scan.^[Note that a different estimand would be the so-called average treatment effect (ATE). For the ATE, the probability contrast is defined for each combination of expert and scan, and then these contrasts are averaged across all experts and scans from the target population [@lundbergWhatYourEstimand2021]].
To complete our definition of the estimand, we have to define our target population that consists of persons, stimuli, and AI-advice: With respect to persons, we are only interested in expert radiologists and medical students at German universities. With respect to stimuli, we are only interested in the head CT scans made from subjects that do or do not suffer from intracerebral hemorrhage. Lastly, we are only interested in AI-advice given by a specific AI-enabled diagnostic decision support system.

Although the estimand is initially defined outside of any statistical model, it is only useful if we find a way to estimate it based on observed data. For our exemplary research question, it is possible to construct an experimental study, where all participants are confronted with the same set of head CT scans, but the kind of AI-advice given for each scan is randomly assigned within participants. This random intervention allows us to produce an empirical estimate of our estimand, although, in reality, each person receives only one kind of AI-advice (correct advice, wrong advice, no advice) for each scan. We will see later how each of the probability expressions in our estimand can be modeled with the same GLMM. Estimating this GLMM based on the data observed in our planned study will produce an estimate for each probability, and these estimates can then be combined to compute an estimate for each of the four probability contrasts. For pedagogical reasons, we will skip the concrete definition of our estimate until we have discussed how to simulate data based on a concrete GLMM in the next section.

## Step 2: Simulate the data generating process

#### THEORY

When the estimand has been defined, the next step in the research process is to write code that simulates the (assumed) data generating process of the planned study. This requires specifying a generative process for all predictor variables used in the final data analysis. While such assumptions can be quite challenging for observational studies or continuous predictor variables, this is less of a problem for experimental studies with only categorical predictor variables. When all predictor variables have been simulated, one can use the structure of a suitable GLMM to simulate the dependent variable. To simulate the GLMM, one requires plausible values for all model parameters. We will discuss strategies on how these values can be obtained later. Because we have full control over the data generating process in a tailored simulation, it is possible to model specific aspects of the planned study, like data missing by design or assuming that some subjects might drop out. The quality of the results of the sample size planning crucially depends on the plausibility of the simulated data generating process. However, we would argue that even a strongly simplified data generating process (e.g. only a small number of interaction effects; only random intercepts and no random slopes; assuming that data is missing completely at random) can yield informative results.

#### PRACTICE

In our case study, we simulate data for an experiment where the diagnostic performance of users of an AI-enabled diagnostic decision support system will be evaluated. Radiologists (task experts) and students/interns (non experts) review a series of head computer tomography (CT) scans to assess the presence of a bleeding. An AI model provides initial diagnostic advice to assist their decision-making. In the control condition, no AI advice is presented. When AI advice is given, this advice can be either correct or incorrect. The type of advice (no advice, wrong advice, correct advice) is randomized within subjects across brain scans. After reviewing a CT scan, participants deliver a medical diagnosis (bleeding or no bleeding), which may be either accurate or inaccurate. This experimental design introduces some missing values by design since the advice is neither correct nor incorrect when no advice is present, which must be taken into account when simulating and analyzing the data. In this example, recruiting task experts (i.e., radiologists) is more challenging due to their limited availability, while non-experts (i.e., students/interns) are more readily accessible. The goal of simulation-based sample size planning is to determine how many task experts and non-experts must be recruited to achieve sufficient statistical power or precision in the planned experiment.

##### Our specific GLMM

In a GLMM, the expected value of the dependent variable $Y$ conditioned on the vector of predictor variables $\mathbf{X}$ and random effects $\mathbf{U}$, transformed by a link function $g()$ is modeled as a linear combination $\eta$ of the predictor variables $\mathbf{X}$, the random effects $\mathbf{U}$ and the model parameters $\mathbf{\beta}$ [@fahrmeirRegressionModelsMethods2021]: 
$$
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta
$$
Equivalently, the conditional expected value is modeled as the linear combination $\eta$, transformed by the inverse link function $g^{-1}()$: 
$$
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u}) = g^{-1}(\eta)
$$
If the dependent variable (i.e., diagnostic decision) $Y$ is a binary variable with values $0$ (i.e., inaccurate), or $1$ (i.e., accurate), the conditional expected value is equivalent to the probability: 
$$
P_{si} := P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
$$
In our case study, $P_{si}$ is the conditional probability that subject $s$ gives the correct response to item (i.e., CT scan) $i$.

In such a setting, we model this probability as
$$
P_{si} = \text{inverse\_logit}(\eta_{si})
$$
with the inverse-logit link $g^{-1}(\eta_{si}) = inverse\_logit(\eta_{si}) = \frac{exp(\eta_{si})}{1 + exp(\eta_{si})}$ or equivalently
$$
\text{logit}(P_{si}) = \eta_{si}
$$
with the logit link $g(P_{si}) = \text{logit}(P_{si}) = \text{ln} (\frac{P_{si}}{1 - P_{si}})$.

In our case study, the probability of making an accurate diagnostic decision is assumed to depend on the predictors:

- $advice\_present_{si}$: whether subject $s$ was presented with AI advice (1) or not (0) when asked to assess item $i$ 
- $advice\_correct_{si}$: whether this advice was correct (1) or not (0) 
- $expert_s$: whether subject $s$ was a task expert (1) or not (0)

and the random effects:

-   $u_{0s}$: the deviation of subject $s$ from the average ability to solve an item (i.e., CT scan) with average difficulty; assumed to be distributed as $u_{0s} \sim N(0, \sigma_S^2)$
-   $u_{0i}$: the deviation of item (i.e., CT scan) $i$ from the average difficulty to be solved by a person with average ability; assumed to be distributed as $u_{0i} \sim N(0, \sigma_I^2)$

In total, we assume the model
$$
\begin{aligned}
\text{logit}[P_{si}] =\ (&\beta_0 + u_{0s} + u_{0i}) + \\
&\beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}
\end{aligned}
$$
or equivalently
$$
\begin{aligned}
P_{si} = \text{inverse\_logit}[&(\beta_0 + u_{0s} + u_{0i}) + \\
&\beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$
with model parameters $\beta_0$, $\beta_e$, $\beta_a$, $\beta_c$, $\beta_{ea}$, $\beta_{ec}$, $\sigma_S$, and $\sigma_I$.

In the GLMM literature, this would be called a binomial GLMM with two random intercepts (for subjects and items), two level-1 predictors ($advice\_present$, $advice\_correct$), one level-2 predictor ($expert$) and two cross-level interactions ($expert \cdot advice\_present$, $expert \cdot advice\_correct$).
To limit complexity, we do not consider random slopes, additional predictors or higher-level interactions.

##### Simulation function in R

The following R function simulates a full dataset structured according to the design of our case study.

```{r}
#| echo: false
b_0 = 0.847
b_e = 1.350
b_a = -1.253
b_c = 2.603
b_ea = 0.790
b_ec = -1.393
```

@@

In the first four lines of the function definition, we set some default parameter values (which we will explain in the next section) and load the packages we use to manipulate and simulate data.^[The faux package [@R-faux] contains useful functions when simulating factorial designs, including random effects.] 
In our case study, each subject (`n_subjects` in total) is assumed to respond to each item (i.e., CT scan; `n_items` in total). Thus, the `add_random` command creates a fully-crossed `data.frame` with `n_subjects` $\times$ `n_items` rows.
We add a between-subject effect with the `add_between` command, simulating that about $25\%$ of subjects are experts.
The next two lines simulate that in $\frac{2}{3}$ of trials, subjects will be presented with AI advice, and if advice is presented, the advice will be correct in about $80\%$ of cases (the variable `advice_correct` is always 0 when no advice is presented). Next, we simulate one random effect for each subject (`u0s`) and for each item (`u0i`).
As assumed by standard GLMMs, the `add_ranef` function draws the random effects from a normal distribution with a mean 0 and a standard deviation specified by the user.
With all design variables done, we are ready to simulate our model equation outlined in the last section.
The linear predictor variable `linpred` ($\eta$ in the GLMM model equations) combines the predictor variables, random effects, and model parameters as assumed by our model.
We then transform the linear predictor with the inverse-link function to compute `y_prob`, the probability that the subject correctly solved the item (in R, the inverse-logit link is computed with `plogis` and the logit link with `qlogis`).
In the final step, we simulate the binary dependent variable `y_bin` (i.e., whether the subject makes an accurate diagnostic decision for the CT scan) by -- for each trial -- drawing from a Bernoulli distribution with success probability `y_prob`.

## Step 3: Specify the population parameters

### THEORY

In the absence of previous studies with the same design or pilot data, researchers require strategies on how to specify the population parameters used in their data simulation. Population parameters are all model parameters estimated in a GLMM, in particular the regression coefficients of the fixed effects and the standard deviation of the random effects (and the correlation between random effects in more complicated models). In contrast to non-hierarchical linear regression, common heuristics based on standardized effect sizes are less useful or not even available for GLMMs. Our strategies to specify population parameters will require access to domain knowledge from domain experts. Because most domain knowledge can only be expressed in unstandardized measurement units of a specific application, we argue that unstandardized effect sizes are usually preferable over standardized effect sizes for tailored simulation-based sample size planning. The basic idea of all strategies is how the data generating process implied by certain values of population parameters can be quantified or visualized in an intuitive way that enables a calibration of population parameters based on the available knowledge of domain experts.

Although we use frequentist model estimation in our tutorial, many strategies described in this chapter are inspired by research on how to monitor the plausibility of model assumptions in applied Bayesian statistics [@gelmanBayesianWorkflow2020].

### PRACTICE

When introducing the simulation function for our case study, we have used theoretically plausible values as defaults for all model parameters ($\beta_0$, $\beta_e$, $\beta_a$, $\beta_c$, $\beta_{ea}$, $\beta_{ec}$, $\sigma_S$, and $\sigma_I$), but have not talked about where these numbers came from.

The starting point for all parameter values in our present case study were based on results from distantly related study designs in the literature. Additionally, we had repeated discussions with our affiliated domain experts in radiology to check whether our assumptions seem plausible.

>>>>

---
title: "Full simulation"
---

::: {.callout-note}
This Notebook is used to run the full simulation for the case study discussed in the manuscript. The simulation takes several hours to run on a Mac Book Pro (13-inch, M1, 2020) with 16 GB memory.
:::

## Simulation functions

@@
